Amplitude metrics for cellular \\ circadian bioluminescence reporters

### 2014-08-29 main_text.tex
Sending to stephanie 2014-08-29

\section*{Introduction}
In mammals, circadian rhythms are endogenous oscillations in gene transcription responsible for coordinating daily changes in physiology.
While the suprachiasmatic nucleus (SCN) in the brain serves as the body's master pacemaker, cells found in peripheral tissues also oscillate in a circadian manner \cite{Lamia2008}.
These peripheral clocks process systemic and SCN-mediated entraining cues, buffering against rhythmic changes in energy availability to maintain metabolic homeostasis \cite{Kornmann2007}.
The amplitude of circadian transcription is a relevant factor, and has been shown to play a critical role in phase resetting and entrainment \cite{Pittendrigh1991, Abraham2010}.
Recent studies have further highlighted the importance of high peripheral clock amplitudes in maintaining metabolic health.
Mice lacking an intact clock have been shown to develop metabolic disease \cite{Marcheva2010}, while low amplitude clock oscillations, whether caused by diet \cite{Hatori2012} or age \cite{Chang2013} have also been tied to metabolic disorders.
An understanding of how clock amplitudes are regulated is therefore a topic of ongoing research with potential therapeutic applications \cite{St.John2014}.
 Unlike the network of oscillators in the SCN, in which intercellular coupling maintains robust amplitudes even in the absence of external cues, peripheral oscillators are thought to lack a direct mechanism to spontaneously synchronize \cite{Welsh2004}.
As a result, populations of peripheral clocks are likely synchronized by common external cues, with stochastic effects and cell heterogeneity driving entrained populations to gradually desynchronize.

@@

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_3.pdf}
    \caption{
{\bfseries Synchrony effects population-level amplitude.}
({\bfseries A}) Stochastic fluctuations cause a population of cells to gradually desynchronize with time.
The phase probability density, $p(\theta, \tilde{t})$, gradually widens as it advances in phase according to the mean period.
({\bfseries B}) The mean amplitude from a population of oscillators is determined by the probability density function $p(\theta, \tilde{t})$ and the oscillator's limit cycle $x^\gamma(\theta)$.
As time passes, population-level rhythms resemble an exponentially damped sinusoid.}
  \end{center}
\end{figure}

@@

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_4.pdf}
    \caption{
{\bfseries Response curves describe different types of amplitude change.}
A population of cells is simulated using a detailed model of circadian rhythms (Model 2).
Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.
({\bfseries A}) The phase response curves are shown for both the single limit cycle oscillator and population average, left.
The population-level ARC is shown together with the single-cell amplitude response, with both curves normalized to $\sigma=1$.
({\bfseries B-C}) Population-level rhythms, $\hat{x}(\tilde{t})$ resulting from perturbations given at the indicated phases in A (red and cyan lines, respectively).
The population transitions from the unperturbed population, $\bar{x}(\tilde{t})$, to the steady-state perturbed population, $\hat{x}_{ss}(\tilde{t})$.
({\bfseries B}) A perturbation at this phase yields a transient amplitude reduction resulting from perturbations at the single-cell level, however, there is little change in population synchrony.
Population-level rhythms are therefore transiently damped before regaining normal amplitudes.
({\bfseries C}) Oscillators are desynchronized, but with a transient increase in limit cycle amplitude.
It therefore takes some time before the amplitude reduction from desynchrony is realized.}
  \end{center}
\end{figure}

@@

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_5.pdf}
    \caption{
{\bfseries Method allows direct comparison between model results and experimental bioluminescence profiles.} 
({\bfseries A}) Response curves for the parameter perturbation used to model the first light pulse in (B).
The initial phase of the system is chosen to coincide with the strong dip in the population-level ARC (left).
({\bfseries B}) De-noised bioluminescence data from Ukai {\itshape et al.} 2007 (top) demonstrates population-level amplitude change resulting from two light pulses \cite{Ukai2007}.
The first pulse desynchronizes the system, while the second resynchronizes.
These perturbations were reproduced {\itshape in silico} (bottom) using the model from Hirota {\itshape et al.} 2012.
The experimental results are qualitatively captured by the model, demonstrating the suitability of the method in capturing bioluminescence experiments.
The relative amplitude of the model equations has been scaled to match that of the normalized bioluminescence profiles.
({\bfseries C}) Example probability density functions for the model trajectory shown in (A).
 The initial population (red) is effectively desynchronized (orange) following the first light pulse.
Following the second light pulse, stronger peaks are seen (teal), which gradually damp as time evolves.} \end{center}
\end{figure}

>>>>

\section*{Introduction}
In mammals, circadian rhythms are endogenous oscillations in gene transcription responsible for coordinating daily changes in physiology.
While the suprachiasmatic nucleus (SCN) in the brain serves as the body's master pacemaker, cells found in peripheral tissues also oscillate in a circadian manner \cite{Lamia2008}.
These peripheral clocks process systemic and SCN-mediated entraining cues, buffering against rhythmic changes in energy availability to maintain metabolic homeostasis \cite{Kornmann2007}.
The amplitude of circadian transcription is a relevant factor, and has been shown to play a critical role in phase resetting and entrainment \cite{Pittendrigh1991, Abraham2010}.
Recent studies have further highlighted the importance of high peripheral clock amplitudes in maintaining metabolic health.
Mice lacking an intact clock have been shown to develop metabolic disease \cite{Marcheva2010}, while low amplitude clock oscillations, whether caused by diet \cite{Hatori2012} or age \cite{Chang2013} have also been tied to metabolic disorders.
An understanding of how clock amplitudes are regulated is therefore a topic of ongoing research with potential therapeutic applications \cite{St.John2014}.
 Unlike the network of oscillators in the SCN, in which intercellular coupling maintains robust amplitudes even in the absence of external cues, peripheral oscillators are thought to lack a direct mechanism to spontaneously synchronize \cite{Welsh2004}.
As a result, populations of peripheral clocks are likely synchronized by common external cues, with stochastic effects and cell heterogeneity driving entrained populations gradually toward desynchrony.

@@

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_3.pdf}
    \caption{
{\bfseries Synchrony affects population-level amplitude.}
({\bfseries A}) Stochastic fluctuations cause a population of cells to gradually desynchronize with time.
The phase probability density, $p(\theta, \tilde{t})$, gradually widens as it advances in phase according to the mean period.
({\bfseries B}) The mean amplitude from a population of oscillators is determined by the probability density function $p(\theta, \tilde{t})$ and the oscillator's limit cycle $x^\gamma(\theta)$.
As time passes, population-level rhythms resemble an exponentially damped sinusoid.}
  \end{center}
\end{figure}

@@

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_4.pdf}
    \caption{
{\bfseries Response curves describe different types of amplitude change.}
A population of cells is simulated using a detailed model of circadian rhythms (Model 2).
Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter ({\itshape Per} transcription rate) for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.
({\bfseries A}) The phase response curves are shown for both the single limit cycle oscillator and population average, left.
The population-level ARC is shown together with the single-cell amplitude response, with both curves normalized to $\sigma=1$.
({\bfseries B-C}) Population-level rhythms, $\hat{x}(\tilde{t})$ resulting from perturbations given at the indicated phases in A (red and cyan lines, respectively).
The population transitions from the unperturbed population, $\bar{x}(\tilde{t})$, to the steady-state perturbed population, $\hat{x}_{ss}(\tilde{t})$.
({\bfseries B}) A perturbation at this phase yields a transient amplitude reduction resulting from perturbations at the single-cell level, however, there is little change in population synchrony.
Population-level rhythms are therefore transiently damped before regaining normal amplitudes.
({\bfseries C}) Oscillators are desynchronized, but with a transient increase in limit cycle amplitude.
It therefore takes some time before the amplitude reduction from desynchrony is realized.}
  \end{center}
\end{figure}

@@

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_5.pdf}
    \caption{
{\bfseries Method allows direct comparison between model results and experimental bioluminescence profiles.} 
({\bfseries A}) Response curves for the parameter perturbation used to model the first light pulse in (B).
The initial phase of the system is chosen to coincide with the strong dip in the population-level ARC (left).
({\bfseries B}) De-noised bioluminescence data from Ukai {\itshape et al.} 2007 (top) demonstrates population-level amplitude change resulting from two light pulses \cite{Ukai2007}.
Circadian oscillations are suppressed by the first light pulse and rescued by the second.
These perturbations were reproduced {\itshape in silico} (bottom) using the model from Hirota {\itshape et al.} 2012.
The experimental results are qualitatively captured by the model, demonstrating the suitability of the method in capturing bioluminescence experiments.
The relative amplitude of the model equations has been scaled to match that of the normalized bioluminescence profiles.
({\bfseries C}) Example probability density functions for the model trajectory shown in (A).
 The initial population (red) is effectively desynchronized (orange) following the first light pulse.
Following the second light pulse, stronger peaks are seen (teal), which gradually damp as time evolves.} \end{center}
\end{figure}
